A Horrocks-type theorem for even orthogonal $K_2$

### 2020-04-27 dl-paper.tex
Better abstract

\begin{document}
\begin{abstract} We prove a $\mathbb{P}^1$-glueing theorem for even-dimensional orthogonal Steinberg groups. This result transfers to the orthogonal case an earlier result of M.~Tulenbaev and is also an analogue for the functor $\K_2$ of a 1982 result of A.~Suslin and V.~Kopeiko. \end{abstract}
\maketitle
\section{Introduction}
Recall that the classical Serre problem on projective modules asks if any projective module over a polynomial ring $R = k[x_1,\ldots, x_n]$ over a field $k$ is free. This problem was positively settled by D.~Quillen~\cite{Qu76} and A.~Suslin~\cite{Su76}, and its solution played an important role in the development of algebraic K-theory. We also refer the reader to the textbook~\cite{Lam10} for a comprehensive account on the problem, its history and the subsequent solution.

After the original Serre problem had been solved, numerous analogous questions drew the attention of specialists (see e.\,g.~\cite{Su77, Su82, Abe83, Tu83, Lam10, St-poly, St-Ded}). For example, A.~Suslin formulated and solved the so-called $K_1$-analogue of Serre's problem. This result asserts that the functor $\K_1(n, R) = \mathrm{GL}_{n}(R)/\E_n(R)$ has the property $\K_1(n, k[x_1, \ldots x_n]) = \K_1(n, k) = k^\times$ for all fields $k$ and $n \geq 3$ see~\cite[Corollary~7.11]{Su77}. Suslin's results were subsequently generalized to $\K_1$-functors modeled on other linear groups (see the definition below). For example, for even-dimensional orthogonal groups the corresponding result was obtained by A.~Suslin and V. Kopeiko in~\cite{Su82}, while for more general types of Chevalley groups of rank $\geq 2$ this is a result of E.~Abe, see~\cite{Abe83}. Recently A.~Stavrova has obtained probably the most general results in this direction: she solved the analogue of Serre problem for the functor $\K_1^G$ modeled on an arbitary isotropic reductive group scheme $G$ of isotropic rank $\geq 2$ over a field (see~\cite[Theorem~1.2]{St-poly}) and also generalized Abe's result to Dedekind domains of arithmetic type (see~\cite[Corollary~1.2]{St-Ded}).

@@

\begin{rem} It is possible to give a different proof of the above lemma in the case $\Phi = \rC_\ell$ for $\ell \geq 5$ under the additional assumption that the characteristic of $k$ is not $2$. Indeed, we have the following chain of isomorphisms:
\[ \K_2(C_\infty, k[X, X^{-1}]) \cong \GW_2^2(k[X,X^{-1}]) \cong \GW_2^2(k) \oplus \GW_1^1(k) \cong \K_2^\mathrm{MW}(k) \oplus \K_1^\mathrm{MW}(k),\]
in which the first two isomorphisms are obtained from~\eqref{GW-concrete} and~\cref{bass-ft} and the last one follows from the isomorphism $\GW_1^1(k) \cong \K_1^\mathrm{MW}(k)$ of~\cite[Lemma~4.1.1]{AF17} and Matsumoto's theorem.
It remains to invoke the stability theorem~\cite[Theorem~9.4]{Pa89} to obtain the assertion in the nonstable case. Thus, we see that the result of Morita and Rehmann agrees with the modern literature on higher Grothendieck--Witt groups. \end{rem}

@@

It is clear that both $G_n$ and $\widetilde{G}_n$ satisfy~\ref{req:left-exact} and that $S_n$ satisfies \ref{req:coeq}.
It is well known that $P_n$ (resp. $\widetilde{P}_n$) is a normal subgroup of $G_n$ (resp. $\widetilde{G}_n$) for $n\geq 3$ (see~\cite{Su82},\cite{Ta84}). Thus,~\ref{req:subfunc} is satisfied for $P_n$ and $\widetilde{P}_n$.

@@

We will also need the following property of relative Steinberg groups which is a special case of a more general property discussed in~\cite[\S~2]{LS17}.
\begin{lemma}\label{lem:lemma32} Let $\Phi$ be a simply-laced root system of rank $\geq 3$,
Consider the following commutative square of canonical homomorphisms.
\begin{equation*} \xymatrix{
    \St(\Phi, B, I) \ar[r] \ar[d] & \St(\Phi, B) \ar[d] \\
    \St(\Phi, R, I) \ar[r] \ar@{-->}^t[ur] & \St(\Phi, R) } \end{equation*}
Then there exists a diagonal homomorphism $t$ which makes the diagram commute.   
\end{lemma} 
\begin{proof}
 Notice that $R$ is isomorphic to the principal localisation of $B$ at $X$
  and that $I$ is uniquely $X$-divisible in the sense of~\cite[\S~4]{LS17}.
 Thus, in the special case $\Phi = \rA_3$ the assertion of the lemma follows from~\cite[Theorem~3]{LS17}.
 In the general case the assertion of the lemma is a corollary of amalgamation theorem~\cite[Theorem~9]{S15}.
\end{proof}

@@

\subsection{Proof of Horrocks theorem.} \label{sec:P1glueing}
Before we proceed with the proof of~\cref{thm:main} let us briefly remind the reader the relevant notation. Recall that $A$ is a local ring with maximal ideal $M$ and $I = M[X, X\inv]$ is an ideal of both $R=A[X, X\inv]$ and $B = A[X\inv] + M[X]$.

>>>>

\begin{document}

\begin{abstract} 
We prove a $\mathbb{P}^1$-glueing theorem for the unstable even-dimensional orthogonal $\K_2$-functor.
This result transfers to the orthogonal case an earlier result of M.~Tulenbaev and is also one of the ingredients needed for the proof of the orthogonal $\K_2$-analogue of Serre's problem. The final result holds for general commutative rings containing $\tfrac{1}{2}$, however some of the intermediate steps in its proof are obtained without this assumption.
\end{abstract}
\maketitle
\section{Introduction}
Recall that the classical Serre problem on projective modules asks if any projective module over a polynomial ring $R = k[x_1,\ldots, x_n]$ over a field $k$ is free. This problem was positively settled by D.~Quillen~\cite{Qu76} and A.~Suslin~\cite{Su76}, and its solution played an important role in the development of algebraic K-theory. We also refer the reader to the textbook~\cite{Lam10} for a comprehensive account on the problem, its history and the subsequent solution.

After the original Serre problem had been solved, numerous analogous questions drew the attention of specialists (see e.\,g.~\cite{Su77, Su82, Abe83, Tu83, Lam10, St-poly, St-Ded}). For example, A.~Suslin formulated and solved the so-called $\K_1$-analogue of Serre's problem. This result asserts that the functor $\K_1(n, R) = \mathrm{GL}_{n}(R)/\E_n(R)$ has the property $\K_1(n, k[x_1, \ldots x_n]) = \K_1(n, k) = k^\times$ for all fields $k$ and $n \geq 3$ see~\cite[Corollary~7.11]{Su77}. Suslin's results were subsequently generalized to $\K_1$-functors modeled on other linear groups (see the definition below). For example, for even-dimensional orthogonal groups the corresponding result was obtained by A.~Suslin and V. Kopeiko in~\cite{Su82}, while for more general types of Chevalley groups of rank $\geq 2$ this is a result of E.~Abe, see~\cite{Abe83}. Recently A.~Stavrova has obtained probably the most general results in this direction: she solved the analogue of Serre problem for the functor $\K_1^G$ modeled on an arbitary isotropic reductive group scheme $G$ of isotropic rank $\geq 2$ over a field (see~\cite[Theorem~1.2]{St-poly}) and also generalized Abe's result to Dedekind domains of arithmetic type (see~\cite[Corollary~1.2]{St-Ded}).

@@

\begin{rem} It is possible to give a different proof of the above lemma in the case $\Phi = \rC_\ell$ for $\ell \geq 5$ under the additional assumption that the characteristic of $k$ is not $2$. Indeed, we have the following chain of isomorphisms:
\[ \K_2(C_\infty, k[X, X^{-1}]) \cong \GW_2^2(k[X,X^{-1}]) \cong \GW_2^2(k) \oplus \GW_1^1(k) \cong \K_2^\mathrm{MW}(k) \oplus \K_1^\mathrm{MW}(k),\]
in which the first two isomorphisms are obtained from~\eqref{GW-concrete} and~\cref{bass-ft} and the last one follows from the isomorphism $\GW_1^1(k) \cong \K_1^\mathrm{MW}(k)$ of~\cite[Lemma~4.1.1]{AF17} and Matsumoto's theorem.
It remains to invoke the stability theorem~\cite[Theorem~9.4]{Pa89} to obtain the assertion in the unstable case. Thus, we see that the result of Morita and Rehmann agrees with the modern literature on higher Grothendieck--Witt groups. \end{rem}

@@

It is clear that both $G_n$ and $\widetilde{G}_n$ satisfy~\ref{req:left-exact} and that $S_n$ satisfies \ref{req:coeq}.
It is well known that $P_n$ (resp. $\widetilde{P}_n$) is a normal subgroup of $G_n$ (resp. $\widetilde{G}_n$) for $n\geq 3$ (see~\cite{Su82},\cite{Ta86}). Thus,~\ref{req:subfunc} is satisfied for $P_n$ and $\widetilde{P}_n$.

@@

We will also need the following property of relative Steinberg groups which is a special case of a more general property discussed in~\cite[\S~2]{LS17}.
\begin{lemma}\label{lem:lemma32} Let $\Phi$ be a simply-laced root system of rank $\geq 3$,
Consider the following commutative square of canonical homomorphisms.
\begin{equation*} \xymatrix{
    \St(\Phi, B, I) \ar[r]^{\mu_B} \ar[d] & \St(\Phi, B) \ar[d] \\
    \St(\Phi, R, I) \ar[r]^{\mu_R} \ar@{-->}^t[ur] & \St(\Phi, R) } \end{equation*}
Then there exists a diagonal homomorphism $t$ which makes the diagram commute.   
\end{lemma} 
\begin{proof}
 Notice that $R$ is isomorphic to the principal localisation of $B$ at $X$
  and that $I$ is uniquely $X$-divisible in the sense of~\cite[\S~4]{LS17}.
 Thus, in the special case $\Phi = \rA_3$ the assertion of the lemma follows from~\cite[Theorem~3]{LS17}.
 In the general case the assertion of the lemma is a corollary of amalgamation theorem~\cite[Theorem~9]{S15}.
\end{proof}

@@

\subsection{Proof of Horrocks theorem.} \label{sec:P1glueing}
Before we proceed with the proof of~\cref{thm:main} let us recall the relevant notation. Recall that $A$ is a local ring with maximal ideal $M$ and $I = M[X, X\inv]$ is an ideal of both $R=A[X, X\inv]$ and $B = A[X\inv] + M[X]$.
